---
title: "Data Science II Final Project Analysis"
author: "Camille Okonkwo"
output:
  pdf_document:
    latex_engine: xelatex
    toc: yes
    toc_depth: 2
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '2'
header-includes:
- \usepackage{fancyhdr}
- \usepackage{lipsum}
- \pagestyle{fancy}
- \fancyhead[R]{\thepage}
- \fancypagestyle{plain}{\pagestyle{fancy}}
editor_options: 
  chunk_output_type: console
--- 
\newpage

```{r setup, include=FALSE}
library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo = T, message = FALSE, results='hide', warning=FALSE}
library(tidymodels)
library(splines)
library(caret)
library(glmnet)
library(table1)
library(kableExtra)
library(summarytools)
library(corrplot)
library(cowplot)
```
\newpage

# Background

A research study aims to identify key factors that predict the severity of COVID-19 illness. This study collects demographic information, clinical variables, and disease severity among participants infected with COVID-19 between 2021 and 2023. The goal is to develop a robust prediction model that can accurately predict COVID-19 severity and understand how predictors impact the risk of severe infection.

# Data

The training data in "severity_training.RData" includes data from 800 participants.

The test data in "severity_test.RData" includes data from another set of 200 participants.

Here is a description of each variable:

* ID (`id`): Participant ID
* Age (`age`): Age
* Gender (`gender`): 1 = Male, 0 = Female
* Race/ethnicity (`race`): 1 = White, 2 = Asian, 3 = Black, 4 = Hispanic
* Smoking (`smoking`): Smoking status; 0 = Never smoked, 1 = Former smoker, 2 = Current smoker
* Height (`height`): Height (in centimeters)
* Weight (`weight`): Weight (in kilograms)
* BMI (`bmi`): Body Mass Index; BMI = weight (in kilograms) / height (in meters) squared
* Hypertension (`hypertension`): 0 = No, 1 = Yes
* Diabetes (`diabetes`): 0 = No, 1 = Yes
* Systolic blood pressure (`SBP`): Systolic blood pressure (in mm/Hg)
* LDL cholesterol (`LDL`): LDL (low-density lipoprotein) cholesterol (in mg/dL)
* Vaccination status at the time of infection (`vaccine`): 0 = Not vaccinated, 1 = Vaccinated
* Depression score (`depression`): Higher scores indicate higher risk for depression
* Severity of COVID-19 infection (`severity`): **Response variable**; 0 = Not severe,  1 = Severe

## Data Preparation
```{r data_partition, echo = T, message = FALSE, results = 'hide', warning=FALSE}
# loading training data
load("data/severity_training.RData") # is depression discrete?

# making discrete variables factors
training_data = training_data |> 
  janitor::clean_names() |> 
  select(-id) |> 
  mutate(gender = as.factor(gender),
         race = as.factor(race),
         smoking = as.factor(smoking),
         hypertension = as.factor(hypertension),
         diabetes = as.factor(diabetes),
         vaccine = as.factor(vaccine),
         severity = as.factor(severity))

# matrix of predictors & vector of response for data set exploration
x_train = model.matrix(severity ~., training_data)[, -1]
y_train = training_data$severity

# loading testing data
load("data/severity_test.RData")

# making discrete variables factors
test_data = test_data |> 
  janitor::clean_names() |> 
  select(-id) |> 
  mutate(gender = as.factor(gender),
         race = as.factor(race),
         smoking = as.factor(smoking),
         hypertension = as.factor(hypertension),
         diabetes = as.factor(diabetes),
         vaccine = as.factor(vaccine),
         severity = as.factor(severity))

# matrix of predictors and vector of response
x_test = model.matrix(severity ~., test_data)[, -1]
y_test = test_data$severity
```
\newpage

# Exploratory analysis and data visualization

## Descriptive Statistics
```{r}
descriptive_table = table1(~ age + gender + race + smoking + height + weight + bmi + hypertension + diabetes + sbp + ldl + vaccine + depression | severity,
                            data = training_data,
                            overall = "Total",
                            caption = "Descriptive Characteristics of Participants, Stratified by Severity of COVID-19 Infection")

ds = t1kable(descriptive_table)
ds
```

## Discrete Variable Visualization
```{r}
# gender x severity
prop_gender <- training_data %>%
  group_by(gender, severity) %>%
  summarise(count = n()) %>%
  mutate(prop = count / sum(count))

gender_labels = c("1" = "Male", "0" = "Female")

ggplot(prop_gender,
       aes(y = factor(gender),
           x = prop,
           fill = factor(severity), label = scales::percent(prop))) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(position = position_dodge(width = 0.9),  vjust = -0.4) +  
  labs(x = "Proportion", y = "Gender", fill = "Severity") +
  ggtitle("Proportion of Severity by Gender") +
  scale_fill_manual(values = c("0" = "blue", "1" = "red")) +
  scale_y_discrete(labels = gender_labels) +
  theme_minimal()

# race x severity 
prop_race <- training_data %>%
  group_by(race, severity) %>%
  summarise(count = n()) %>%
  mutate(prop = count / sum(count))

race_labels = c("1" = "White", "2" = "Asian", "3" = "Black", "4" = "Hispanic")

ggplot(prop_race,
       aes(y = factor(race),
           x = prop,
           fill = factor(severity), label = scales::percent(prop))) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(position = position_dodge(width = 0.9),  vjust = -0.4) +  
  labs(x = "Proportion", y = "Race", fill = "Severity") +
  ggtitle("Proportion of Severity by Race") +
  scale_fill_manual(values = c("0" = "blue", "1" = "red")) +
  scale_y_discrete(labels = race_labels) +
  theme_minimal()

# smoking status x severity
prop_smoking <- training_data %>%
  group_by(smoking, severity) %>%
  summarise(count = n()) %>%
  mutate(prop = count / sum(count))

smoking_labels = c("1" = "Former Smoker", "2" = "Current Smoker", "0" = "Never Smoked")

ggplot(prop_smoking,
       aes(y = factor(smoking),
           x = prop,
           fill = factor(severity), label = scales::percent(prop))) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(position = position_dodge(width = 0.9),  vjust = -0.4) +  
  labs(x = "Proportion", y = "Smoking Status", fill = "Severity") +
  ggtitle("Proportion of Severity by Smoking Status") +
  scale_fill_manual(values = c("0" = "blue", "1" = "red")) +
  scale_y_discrete(labels = smoking_labels) +
  theme_minimal()

# hypertension status x severity
prop_hyp <- training_data %>%
  group_by(hypertension, severity) %>%
  summarise(count = n()) %>%
  mutate(prop = count / sum(count))

hyp_labels = c("0" = "No", "1" = "Yes")

ggplot(prop_hyp,
       aes(y = factor(hypertension),
           x = prop,
           fill = factor(severity), label = scales::percent(prop))) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(position = position_dodge(width = 0.9),  vjust = -0.4) +  
  labs(x = "Proportion", y = "Hypertension Status", fill = "Severity") +
  ggtitle("Proportion of Severity by Hypertension Status") +
  scale_fill_manual(values = c("0" = "blue", "1" = "red")) +
  scale_y_discrete(labels = hyp_labels) +
  theme_minimal()

# diabetes x severity
prop_dia <- training_data %>%
  group_by(diabetes, severity) %>%
  summarise(count = n()) %>%
  mutate(prop = count / sum(count))

dia_labels = c("0" = "No", "1" = "Yes")

ggplot(prop_dia,
       aes(y = factor(diabetes),
           x = prop,
           fill = factor(severity), label = scales::percent(prop))) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(position = position_dodge(width = 0.9),  vjust = -0.4) +  
  labs(x = "Proportion", y = "Diabetes Status", fill = "Severity") +
  ggtitle("Proportion of Severity by Diabetes Status") +
  scale_fill_manual(values = c("0" = "blue", "1" = "red")) +
  scale_y_discrete(labels = dia_labels) +
  theme_minimal()

# vaccine x severity
prop_vaccine <- training_data %>%
  group_by(vaccine, severity) %>%
  summarise(count = n()) %>%
  mutate(prop = count / sum(count))

vax_labels = c("0" = "Not vaccinated", "1" = "Vaccinated")

ggplot(prop_vaccine,
       aes(y = factor(vaccine),
           x = prop,
           fill = factor(severity), label = scales::percent(prop))) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(position = position_dodge(width = 0.9),  vjust = -0.4) +  
  labs(x = "Proportion", y = "Vaccination Status at Time of Infection", fill = "Severity") +
  ggtitle("Proportion of Severity by Vaccination Status") +
  scale_fill_manual(values = c("0" = "blue", "1" = "red")) +
  scale_y_discrete(labels = vax_labels) +
  theme_minimal()
```

## Continuous Variable Visualization
```{r}
theme1 <- trellis.par.get()
theme1$plot.symbol$col <- rgb(.2, .4, .2, .5)
theme1$plot.symbol$pch <- 2
theme1$plot.line$col <- rgb(.8, .1, .1, 1)
theme1$plot.line$lwd <- 2
theme1$strip.background$col <- rgb(.0, .2, .6, .2)
trellis.par.set(theme1)

featurePlot(
  x_train[, -c(2, 3, 4, 5, 6, 7, 11, 12, 15)],
  y_train,
  plot = "box",
  labels = c("", "Y"),
  type = c("p", "smooth"),
  layout = c(4, 3))
```

\newpage

# Model training
```{r}
```
\newpage

# Results
```{r}

```
\newpage

# Conclusion
```{r}
```
\newpage


